Quantifying the Relationship Between Genetic Diversity and Population Size Suggests Natural Selection Cannot Explain Lewontin's Paradox

### 2021-08-09 manuscript/elife/elife-manuscript.tex
more ms reviewer feedback

Since the phylogenetic mixed-effects model simultaneously estimates the
variance of the phylogenetic effect ($\sigma_p^2$) and the residual variance
($\sigma_r^2$), these can be used to estimate the phylogenetic signal, $\lambda
= \nicefrac{\sigma_p^2}{(\sigma_p^2 + \sigma_r^2)}$
(\cite{Lynch1991-tp,De_Villemereuil2014-kt}; see \cite{Freckleton2002-ly} for a
comparison to Pagel's $\lambda$). If the residuals were free of correlations
due to shared phylogenetic history, then $\lambda = 0$ and all the variance
could be explained by evolution on the tips; this is analogous to Lynch's
conjecture that coalescent times should be free of phylogenetic signal
(\citeyear{Lynch2011-qv}). In the relationship between population size and
diversity, the posterior mean of $\lambda = 0.67$ (90\% credible interval
$[0.58, 0.75]$) indicates a majority of the variance perhaps might be
due to shared phylogenetic history (\FIG{pcm}B).

@@

Interestingly, the diversity node-height test reveals two rate shifts at deeper
splits (\FIG{pcm}C, top left) around 570 Mya. These nodes represent the
branches between tunicates and vertebrates in chordates, and cephalopods and
pleistomollusca (bivalves and gastropods) in molluscs.  While the
cephalopod-pleistomollusca split outlier may be an artifact of having a single
cephalopod (\emph{Sepia officinalis}) in the phylogeny, the tunicate-vertebrate
split outlier is driven by the low diversity of vertebrates and the
previously-documented exceptionally high diversity of tunicates (sea squirts;
\cite{Nydam2010-kg,Small2007-mt}). This deep node representing a rate shift in
diversity could reflect a change in either effective population size or
mutation rate, and there is some evidence of both in this genus \emph{Ciona}
\citep{Small2007-mt,Tsagkogeorga2012-lv}. Neither of these deep rate shifts in
diversity is mirrored in the population size node-height test (Figure
\FIG{pcm}C, top right). Rather, it appears a trait impacting diversity but not
census size (e.g. mutation rate or offspring distributions) has experienced a
shift on the lineage separating tunicates and vertebrates. At nearly 600 Mya,
these deep nodes illustrate that expected effective population sizes (and thus
coalescent times) can share phylogenetic history, due to phylogenetic inertia
in some combination of population size, reproductive system, and mutation
rates.

@@

Then, I predict the expected diversity ($\pi_{BGS+HH}$) under background
selection and hitchhiking, assuming $N_e= N_c$ and that all species had the
rate of sweeps and strength of BGS as \emph{D. melanogaster}. Since neutral
mutation rates $\mu$ are unknown and vary across species, I calculate the range
of predicted $\pi_{BGS+HH}$ estimates for $\mu = 10^{-8}$--$10^{-9}$ (using the
four-alleles model, \cite{Tajima1996-rb}), and compare this to the observed
relationship between $\pi$ and $N_c$ in \FIG{linksel}B.  Under these parameters
and assumptions, linked selection begins to appreciably constrain diversity for
$N_c \gtrsim 10^7$, since $S(\gamma_\text{Dmel}, J_\text{Dmel}, L) \approx
10^{-7}$--$10^{-8}$ and linked selection dominates drift when $S(\gamma, J, L)
> \nicefrac{1}{2N}$. Overall, this reveals two problems for the hypothesis that
linked selection could solve Lewontin's Paradox. First, low to mid-$N_c$
species (census sizes between $10^{4}$--$10^{7}$) have sufficiently long map
lengths that their diversity levels are only moderately reduced by linked
selection, leading to a wide gap between predicted and observed diversity
levels. For this not to be the case, the rate of adaptive mutations or the
deleterious mutation rate would need to be orders of magnitude higher for
species within this range than in \emph{Drosophila melanogaster}, which is
incompatible with the rate of adaptive protein substitutions across species
\citep{Galtier2016-dq} and overall mutation rates \citep{Lynch2010-ki}.
Furthermore, linked selection has been quantified in humans, which fall in this
census size range, and has been found to be relatively weak
\citep{McVicker2009-ax,Hernandez2011-gs,Hellmann2008-ic,Cai2009-by,Boyko2008-tj}.
Second, while hitchhiking and BGS can reduce predicted diversity levels for
high-$N_c$ species ($N_c > 10^{12}$) to observed levels, this would imply
available estimates of $\pi_0$ are underestimated by several orders of
magnitude in \emph{Drosophila} (\FIGSUPP[linksel]{reductions}B). The high
reductions in $\pi$ predicted here (compared to those of
\cite{Elyashiv2016-vt}) are a result of using $N_c$, rather than $N_e =
\nicefrac{\pi_0}{4 \mu}$ in the denominator of Equation \eqref{eqn:div}, which
leads to a very high rate of sweeps in the population.  I do not consider
selective interference, though the saturation of adaptive substitutions per
Morgan would only act to limit the reduction in diversity
\citep{Weissman2012-aa}, and thus these results are conservative.  Overall,
while linked selection could decouple diversity from census size for high-$N_c$
species, recurrent hitchhiking and background selection seem unlikely to
explain the observed patterns of diversity across species under our
understanding of the range of parameter estimates.

@@

One limitation of the dataset in this study is that diversity estimates are
collated from a variety of sources rather than estimated with a single
bioinformatic pipeline. This leads to technical noise across diversity
estimates; perhaps the relationship between $\pi$ and $N_c$ found here could be
tighter with a standardized bioinformatic pipeline. In addition to this
technical variation, there might be systematic bioinformatic sources of bias in
diversity estimates. For example high-diversity sequences may fail to align to
the reference genome and end up unaccounted for, leading to a downward bias.
Alternatively, high-diversity sequences might map to the reference genome, but
adjacent mis-matching SNPs might be mistaken for a short insertion or deletion.
While these issues might adversely affect the estimates in high-diversity
species, it is unlikely they will qualitatively change the observed
$\pi$--$N_c$ relationship.

@@

Lewontin's Paradox arises from a comparison of diversity across species, yet it
has been disputed whether such comparisons require phylogenetic comparative
methods. Extending previous work that has accounted for phylogeny in particular
clades \citep{Leffler2012-zj}, or using taxonomical-level averages
\citep{Romiguier2014-bp}, I show that the positive relationship between
diversity and census size is significant using a mixed-effects model with a
time-calibrated phylogeny. Additionally, I find a high degree of phylogenetic
signal, evidence of deep shifts in the rate of evolution of genetic diversity,
and that arthropods and chordates form clusters. Overall, this suggests that
previous concerns about phylogenetic non-independence in comparative population
genetic studies were warranted \citep{Gillespie1991-qa,Whitney2010-ud}.
Notably, Lynch (2011) has argued that PCMs for pairwise diversity are
unnecessary, since mutation rate evolution is fast and thus free of
phylogenetic inertia, sampling variance should exceed the variance due to
phylogenetic shared history, and coalescent times are much less than divergence
times. Since my findings suggest PCMs are necessary in some cases, it is
worthwhile to address these points.

First, Lynch has correctly pointed out that while coalescent times are much
less than divergence times and should be free of phylogenetic shared history,
the factors that determine coalescent times (e.g.  mutation rates and effective
population size) may not be (2011). In other words, coalescent times are free
from phylogenetic shared history \emph{were we to condition} on these causal
factors that could be affected by shared phylogenetic history. My estimates of
phylogenetic signal in diversity, by contrast, are not conditioned on these
factors. Importantly, even ``correcting for" phylogeny implicitly favors
certain causal interpretations over others
\citep{Westoby1995-wz,Uyeda2018-wf}. Future work could try to untangle what
causal factors determine coalescent times across species, as well as how these
factors evolve across macroevolutionary timescales. Second, it is a
misconception that a fast rate of trait evolution necessarily reduces
phylogenetic signal \citep{Revell2008-in}, and that if either or both
variables in a regression are free of phylogenetic signal, PCMs are unnecessary
\citep{Revell2010-mf,Uyeda2018-wf}. The evidence of high phylogenetic
signal found in this study suggests PCMs are needed, in part to avoid spurious
results from phylogenetic pseudoreplication.

>>>>

Since the phylogenetic mixed-effects model simultaneously estimates the
variance of the phylogenetic effect ($\sigma_p^2$) and the residual variance
($\sigma_r^2$), these can be used to estimate the phylogenetic signal, $\lambda
= \nicefrac{\sigma_p^2}{(\sigma_p^2 + \sigma_r^2)}$
(\cite{Lynch1991-tp,De_Villemereuil2014-kt}; see \cite{Freckleton2002-ly} for a
comparison to Pagel's $\lambda$). If the residuals were free of correlations
due to shared phylogenetic history, then $\lambda = 0$ and all the variance
could be explained by evolution on the tips; this is analogous to Lynch's
conjecture that coalescence times should be free of phylogenetic signal
(\citeyear{Lynch2011-qv}). In the relationship between population size and
diversity, the posterior mean of $\lambda = 0.67$ (90\% credible interval
$[0.58, 0.75]$) indicates a majority of the variance perhaps might be
due to shared phylogenetic history (\FIG{pcm}B).

@@

Interestingly, the diversity node-height test reveals two rate shifts at deeper
splits (\FIG{pcm}C, top left) around 570 Mya. These nodes represent the
branches between tunicates and vertebrates in chordates, and cephalopods and
pleistomollusca (bivalves and gastropods) in molluscs.  While the
cephalopod-pleistomollusca split outlier may be an artifact of having a single
cephalopod (\emph{Sepia officinalis}) in the phylogeny, the tunicate-vertebrate
split outlier is driven by the low diversity of vertebrates and the
previously-documented exceptionally high diversity of tunicates (sea squirts;
\cite{Nydam2010-kg,Small2007-mt}). This deep node representing a rate shift in
diversity could reflect a change in either effective population size or
mutation rate, and there is some evidence of both in this genus \emph{Ciona}
\citep{Small2007-mt,Tsagkogeorga2012-lv}. Neither of these deep rate shifts in
diversity is mirrored in the population size node-height test (Figure
\FIG{pcm}C, top right). Rather, it appears a trait impacting diversity but not
census size (e.g. mutation rate or offspring distributions) has experienced a
shift on the lineage separating tunicates and vertebrates. At nearly 600 Mya,
these deep nodes illustrate that expected effective population sizes (and thus
coalescence times) can share phylogenetic history, due to phylogenetic inertia
in some combination of population size, reproductive system, and mutation
rates.

@@

Then, I predict the expected diversity ($\pi_{BGS+HH}$) under background
selection and hitchhiking, assuming $N_e= N_c$ and that all species had the
rate of sweeps and strength of BGS as \emph{D. melanogaster}. Since neutral
mutation rates $\mu$ are unknown and vary across species, I calculate the range
of predicted $\pi_{BGS+HH}$ estimates for $\mu = 10^{-8}$--$10^{-9}$ (using the
four-alleles model, \cite{Tajima1996-rb}), and compare this to the observed
relationship between $\pi$ and $N_c$ in \FIG{linksel}B.  Under these parameters
and assumptions, linked selection begins to appreciably constrain diversity for
$N_c \gtrsim 10^7$, since $S(\gamma_\text{Dmel}, J_\text{Dmel}, L) \approx
10^{-7}$--$10^{-8}$ and linked selection dominates drift when $S(\gamma, J, L)
> \nicefrac{1}{2N}$. Overall, this reveals two problems for the hypothesis that
linked selection could solve Lewontin's Paradox. First, low to mid-$N_c$
species (census sizes between $10^{4}$--$10^{7}$) have sufficiently long map
lengths that their diversity levels are only moderately reduced by linked
selection, leading to a wide gap between predicted and observed diversity
levels. For this not to be the case, the rate of adaptive mutations or the
deleterious mutation rate would need to be orders of magnitude higher for
species within this range than in \emph{Drosophila melanogaster}, which is
incompatible with the rate of adaptive protein substitutions across species
\citep{Galtier2016-dq} and overall mutation rates \citep{Lynch2010-ki}.
Furthermore, linked selection has been quantified in humans, which fall in this
census size range, and has been found to be relatively weak
\citep{McVicker2009-ax,Hernandez2011-gs,Hellmann2008-ic,Cai2009-by,Boyko2008-tj}.
Second, while hitchhiking and BGS can reduce predicted diversity levels for
high-$N_c$ species ($N_c > 10^{12}$) to observed levels, this would imply
available estimates of $\pi_0$ are underestimated by several orders of
magnitude in \emph{Drosophila} (\FIGSUPP[linksel]{reductions}B). The high
reductions in $\pi$ predicted here (compared to those of
\cite{Elyashiv2016-vt}) are a result of using $N_c$, rather than $N_e =
\nicefrac{\pi_0}{4 \mu}$ in the denominator of Equation \eqref{eqn:div}, which
leads to a very high rate of sweeps in the population. I do not consider
selective interference, though the saturation of adaptive substitutions per
Morgan would only act to limit the reduction in diversity
\citep{Weissman2012-aa}, and thus these results are conservative.

Finally, the poor fit between observed and predicted levels of diversity across
species is not remedied by stronger selection parameters. In
\FIGSUPP[linksel]{linked-sel-alt}B, I increase both selection parameters $U$
and $\gamma$ ten-fold each, and find the same qualitative pattern: on a log-log
scale the relationship between $N_c$ and $\pi$ is linear, while the predicted
diversity under linked selection is non-linear with $N_c$. Under this ten-fold
higher selection regime, there is more overlap between observed and predicted
levels of diversity, but diversity is severely \emph{under-predicted} for
high-$N_c$ species. Additionally, this would imply that selection in
low-to-mid-$N_c$ species is ten-folder higher than estimated in
\emph{Drosophila melanogaster}, which is implausible.

Overall, predicted levels of diversity under models of recurrent hitchhiking
and background selection fit the observed relationship between $N_c$ and $\pi$
poorly, and this is not overly sensitive to parameter choice. These findings
extend the argument of \cite{Coop2016-gx}, showing that not only are previous
empirical estimates of the impact of linked selection unable to explain
Lewontin's Paradox, but it is unlikely current linked selection models under a
wide range of selection parameters are capable of explaining the observed
patterns of diversity across species.

@@

One limitation of this study is that diversity estimates are collated from a
variety of sources rather than estimated with a single bioinformatic pipeline.
This leads to technical noise across diversity estimates; perhaps the
relationship between $\pi$ and $N_c$ found here could be tighter with a
standardized bioinformatic pipeline. In addition, there might be systematic
bioinformatic sources of bias: for example high-diversity sequences may fail to
align to the reference genome and end up unaccounted for, leading to a downward
bias.  Alternatively, a high-diversity sequences might map to the reference
genome, but adjacent mis-matching SNPs might be mistaken for a short insertion
or deletion.  While these issues might affect estimates in high-diversity
species, it is unlikely to change the qualitative relationship between
$\pi$--$N_c$.

@@

Lewontin's Paradox arises from a comparison of diversity across species, yet it
has been disputed whether such comparisons require phylogenetic comparative
methods. Extending previous work that has accounted for phylogeny in particular
clades \citep{Leffler2012-zj}, or using taxonomical-level averages
\citep{Romiguier2014-bp}, I show that the positive relationship between
diversity and census size is significant using a mixed-effects model with a
time-calibrated phylogeny. Additionally, I find a high degree of phylogenetic
signal, evidence of deep shifts in the rate of evolution of genetic diversity,
and that arthropods and chordates form clusters. Overall, this suggests that
previous concerns about phylogenetic non-independence in comparative population
genetic studies were warranted \citep{Gillespie1991-qa,Whitney2010-ud}.
Notably, Lynch (2011) has argued that PCMs for pairwise diversity are
unnecessary, since mutation rate evolution is fast and thus free of
phylogenetic inertia, sampling variance should exceed the variance due to
phylogenetic shared history, and coalescence times are much shorter than
divergence times. Since my findings suggest PCMs are necessary in some cases,
it is worthwhile to address these points.

First, Lynch has correctly pointed out that while coalescence times are much
less than divergence times and should be free of phylogenetic shared history,
the factors that determine coalescence times (e.g.  mutation rates and
effective population size) may not be (2011). In other words, coalescence times
are free from phylogenetic shared history \emph{were we to condition} on these
causal factors that could be affected by shared phylogenetic history. My
estimates of phylogenetic signal in diversity, by contrast, are not conditioned
on these factors. Importantly, even ``correcting for" phylogeny implicitly
favors certain causal interpretations over others
\citep{Westoby1995-wz,Uyeda2018-wf}.  Future work could try to untangle what
causal factors determine coalescence times across species, as well as how these
factors evolve across macroevolutionary timescales. Second, it is a
misconception that a fast rate of trait evolution necessarily reduces
phylogenetic signal \citep{Revell2008-in}, and that if either or both variables
in a regression are free of phylogenetic signal, PCMs are unnecessary
\citep{Revell2010-mf,Uyeda2018-wf}. The evidence of high phylogenetic signal
found in this study suggests PCMs are necessary when fitting the relationship
between $N_c$ and $\pi$ in order to account for correlated residuals among
closely-related species, and to avoid spurious results from phylogenetic
pseudoreplication.
